Article title

### 2018-10-01 docs/rupee.tex
update plots

\usepackage{lmodern}

@@

@@

\section{Introduction}

Proteins represent the functional end-product within the central dogma of molecular biology \citep{Crick1970}.
As such, understanding protein structure is a central goal within structural bioinformatics. 
Protein structure determination, prediction, alignment, and search all serve to advance this understanding. 
Below, we present our new approach to a fast, scalable, and purely geometric protein structure search we refer to with the cumbersome but catchy acronym of \emph{RUn Position Encoded Encodings} of residue descriptors (RUPEE).

For comparisons, we examine results from the CATHEDRAL structural scan \citep{Redfern2007} available at the CATH website \citep{Sillitoe2015}.
In addition to recognizing domains in multidomain proteins, CATHEDRAL also can be used to find structural neighbors among CATH domains for an uploaded protein data bank (PDB) file, that is, where an identifier is not provided that can be used for retrieval of pre-calculated results. 
CATHEDRAL is effective at this task but can take upwards of 10 minutes to produce results against CATH s35 representatives.

Restricting our initial approach to protein domains, we have indexed CATH v4.1 domains, consisting of 308,999 structures, and developed a feature similar to CATHEDRAL that performs a protein structure search based on a PDB file, returning results in seconds rather than minutes.
To evaluate RUPEE, we have made comparisons to the output of CATHEDRAL along two dimensions, quality of results and response times. 
In both cases, RUPEE performs better on average for the domains we evaluated.

@@

For the remainder of this paper, we first discuss some related work to provide context for our approach followed by a description of our method. 
We end with the results of evaluation with CATHEDRAL along with an analysis and conclusion.

@@

@@

Broadly, we define a linear encoding of protein structure and convert this linear encoding into a bag of features amenable to big data processing.
Protein structure searches that use linear encodings are not unique \citep{Carpentier2005,Daniluk2011,Ritchie2012}.
The novelty of our approach lies in its remarkable performance given its simplicity. 
Additionally, elements of our approach can be isolated and found to be useful in their own right.

@@

\begin{figure}[t]
\begin{center}
\includegraphics{ramachandran}
\end{center}
\caption{Ramachandran plot of randomly sampled torsion angles}
\label{fig:ramachandran}
\end{figure}

@@

Despite their utility and familiarity, Ramachandran plots \citep{Ramachandran1968} represent angular data using a square plot better suited for scalar data.
This leads to the unwieldy arrangement where the top part of the plot is continuous with the bottom and the left is continuous with the right.

To identify regions of torsion angles, we randomly sampled 10,000 residues from high-resolution, CATH s35 representatives to account for precision and redundancy, respectively. 
A Ramachandran plot of the sampled torsions angles is shown in \figurename~\ref{fig:ramachandran}. 
As can be seen, a single cluster of residues, consisting primarily of $\beta$-strands, appears at all 4 corners of the Ramachandran plot.

This continuity problem was partially addressed in \citet{Karplus2010} using \emph{wrapped} and \emph{mirrored} plots. 
Both wrapped and mirrored plots take advantage of the sparsely populated areas of the Ramachandran plot at $\phi = 0\degree$ and $\psi = -120\degree$.
However, with larger samples of torsion angles, the area at $\psi = -120\degree$ becomes less sparse. 
The use of a polar plot resolves this elegantly by only requiring one break in continuity at $\phi = 0\degree$.

\figurename~\ref{fig:torsion} plots the same torsion angles appearing in \figurename~\ref{fig:ramachandran} using a polar plot. 
In this plot, $\phi$ corresponds to the radius $r$ and $\psi$ corresponds to the angle $\theta$ in traditional polar plots. 
Notice the residues appearing at the 4 corners of the Ramachandran plot now appear in one continuous region of the polar plot centered at $\phi = \pm180\degree$ and $\psi = \pm180\degree$.

\begin{figure}[t]
\begin{center}
\includegraphics{torsion}
\end{center}
\caption{Polar plot of randomly sampled torsion angles}
\label{fig:torsion}
\end{figure}

@@

The polar plot described above is used to define torsion angle regions for each secondary structure assignment. 
The eight DSSP secondary structure assignment codes defined in \citet{Kabsch1983} divide into three groups in which torsion angle regions are roughly the same: `G',`H',`I', and `T' corresponding to $3_{10}$-helix, $\alpha$-helix, $\pi$-helix, and turn, respectively; `E' and `B' corresponding to $\beta$-strand and $\beta$-bridge, respectively; and `S' and `C' corresponding to bend and coil, respectively.

@@

%\begin{figure}[t]
%\begin{center}
%\includegraphics{torsion_refs}
%\end{center}
%\caption{Polar plots of randomly sampled torsion angles with designated descriptors for region and DSSP code combinations}
%\label{fig:torsion_refs}
%\end{figure}

@@

\begin{table}[!t]
    \caption{DSSP Code/Descriptor Map}
    \label{tab:codes}
\centering
\begin{tabular}{lll}
\toprule
    \head{DSSP Codes} & \head{SS} & \head{Descriptors}\\
\midrule
    G, H, I & helix & 1-4\\
    E & strand & 5-7\\
    S, C & bend, coil & 8-10\\
    T & turn & 11\\
    B & bridge & 12\\
\bottomrule
\end{tabular}
\end{table}

Our initial use of Ramachandran plots to identify regions resulted in the frequent case of highly similar domains from the same CATH s95 clusters being considered significantly different. 
Alignment of these domains in PyMOL \citep{PyMOL} and the development of a script to label residues with their descriptors showed that slight differences in structure often resulted in distinct descriptors being assigned to aligned residues. 
Once we defined regions using the less confusing polar plots, this discrepancy was no longer observed.

@@

Secondary structure assignments for termini are not always consistent among the variety of secondary structure assignment tools \citep{Martin2005}. 
We also observed this to be the case for DSSP assignments across homologous domains. 
Here, we introduce a heuristic we call \emph{secondary structure extension} (SSE) to address this variance.

@@

Though \citet{Martin2005} lends some support to this heuristic, we use it primarily based on its performance.

@@

Once a linear encoding for a protein structure is obtained, it needs to be further transformed into a representation suitable for fast and scalable similarity comparisons to other structures.
The processing of text documents within Information Retrieval~(IR) has long been used to satisfy these requirements using bag representations.
There are two distinct categories of representations for documents, syntactic and semantic, and much of the research applying IR to protein structure search has focused on the latter \citep{Aungand2004,Zhang2010,Budowski2010}.

We have adapted the syntactic approach to document similarity, often referred to as shingling \citep{Broder1997a}, to our linear encoding of protein structure. 
We transform a linear sequence of descriptors into a multiset of shingles consisting of 4 consecutive descriptors.
The overlap between shingles ensures some of the order information within the original sequence is preserved in the bag.

@@

Next, each shingle $s$ is hashed to an integer as shown in~(\ref{E:hashdef}). 
The hash function used is a simplification of the hash function used in the Rabin-Karp algorithm \citep{Karp1987}.
The prime number $13$ is used as the base since it is large enough to spread the descriptor values out in hash space without collisions. 
\begin{gather}\label{E:hashdef}
    s_{hash} = s_1 \times 13^3 + s_2 \times 13^2 + s_3 \times 13 + s_4
\end{gather}
Subsequently, the multiset in (\ref{E:shinglebag}) becomes the following bag of integers.
\begin{align}\label{E:hashbag}
    \begin{split}
    \{\,&11900, 11900, 11906, 11984, 12992, \\
        &26096, 25082, 11900, 11900, 11900, \\
        &11900, 11905, 11969 \,\}
    \end{split}
\end{align}

@@

Now that we have a representation of a protein structure as a bag of integers, similarity between any two structures $a$ and $b$ is defined as the Jaccard similarity \citep{Levan1971} for multisets,
\begin{align}
    J(a,b) = \frac{\sum_i min(a_i, b_i)}{\sum_i max(a_i,b_i)}\text{,}
\end{align}
where $i$ ranges over all possible shingle hashes $s_i$ and $a_i$ and $b_i$ give the counts of shingle hash $s_i$ in structures $a$ and $b$, respectively.

@@

In IR, the bag of shingles representation of documents is used in the near dupe clustering of documents \citep{Broder1997b}.
One application of near dupe clustering is in the review stage of Electronic-Discovery \citep{Joshi2008}, which is the most expensive stage in a discovery process. 
Often millions of documents must be examined by a staff of attorneys to make a reasonable effort at providing all documents relevant to the discovery request. 
Grouping documents into near dupe clusters and assigning all documents within a cluster to a single reviewer reduces duplication of effort.

In the case of near dupe clustering, each document must be compared to every other document in the collection, taking quadratic time.
For this task, min-hashing \citep{Broder1998} and locality sensitive hashing (LSH) \citep{Indyk1998} can be combined to reduce this to subquadratic time. 
Although we do not near dupe cluster domains, we can still leverage the techniques of min-hashing and LSH to speed up protein structure search by a large constant factor.

Min-hashing is used to randomly select items from a set of items by repeatedly randomly hashing the items, sorting the hashes into a list, and then selecting the minimum item in each permuted list.
If the same random permutation of items is performed on each set of items in a collection, the key result is that the probability of matching min-hashes is equal to the Jaccard similarity \citep{Broder1998}.
In order to approximate the Jaccard similarity for a given pair of sets, a sufficient number of min-hashes must be obtained.

In our case, the items are bags of shingle hashes for protein structures from which we obtain 60 min-hashes as described in \citet{Rajaraman2012}. 
Given the key result above, the Jaccard similarity can now be approximated by the proportion of matching min-hashes.

Weighted min-hashing \citep{Chum2008} with integral weights is an extension to min-hashing that can be applied to multisets where the frequency of a hash within a multiset is used as its weight.

Next, we use the LSH banding technique as described in \citet{Rajaraman2012}.
The key result of the banding technique is that if \emph{any} band positions are a match for a given pair of structures, the probability that a specific similarity threshold has been met can be calculated. 
We use 20 bands of 3-min-hashes where the probability of a Jaccard similarity of 60\% or greater is approximately 99\%. 
Banding allows the problem of finding similar items to be parallelized across bands since all that is needed for a match is a single band match.

>>>>

\usepackage{textcomp}
\usepackage{lmodern}
\usepackage{gensymb}
\usepackage{booktabs}

@@

\hbadness=99999

\newcommand{\head}[1]{\textnormal{\textbf{#1}}}

@@

\section{INTRODUCTION}

Proteins represent the functional end-product within the central dogma of molecular biology \cite{Crick1970}.
As such, understanding protein structure is a central goal within structural bioinformatics. 
Protein structure determination, prediction, alignment, and search all serve to advance this understanding. 
Below, we present our new approach to a fast, scalable, and purely geometric protein structure search we refer to with the acronym of \emph{RUn Position Encoded Encodings} of residue descriptors (RUPEE).

%For comparisons, we examine results from the mTM-align structure search \cite{Dong2018}, the CATHEDRAL structural scan \cite{Redfern2007} available at the CATH website \cite{Sillitoe2015}, and the secondary structure matching (SSM) search \cite{Krissinel2004}.
%To evaluate RUPEE, we have made comparisons to the output of the above tools along two dimensions, quality of results and response times. 
%In the case CATHEDRAL and SSM, RUPEE performs better on average for the domains we evaluated.

%Restricting our initial approach to protein domains, we have indexed CATH v4.1 domains, consisting of 308,999 structures, and developed a feature similar to CATHEDRAL that performs a protein structure search based on a PDB file, returning results in seconds rather than minutes.

@@

We first discuss some related work to provide a context for our approach followed by a description of our method. 
We end with a comparison of results against the mTM-align structure search \cite{Dong2018}, the secondary structure matching (SSM) search \cite{Krissinel2004}, and the CATHEDRAL structural scan \cite{Redfern2007} available at the CATH website \cite{Sillitoe2015}.

@@

\section{RELATED WORK}

Pairwise alignment involves finding a set of spatial rotations and translations for two protein structures that minimizes a distance metric. 
Most commonly, the root mean squared deviation (RMSD) between \cas of aligned residues is minimized.

The typical use case of aligning one protein structure to another does not impose tight response time requirements. 
For this reason, pairwise alignments can focus on accuracy. 
On the other hand, a protein structure search can involve thousands of comparisons and accuracy is often balanced against speed. 
In this case, pairwise alignment is still useful for evaluating the results of a search, and this is the approach we take.

For pairwise alignment, Combinatorial Extensions (CE) \citep{Shindyalov1998} and FATCAT \citep{Ye2003} are among the most popular tools, representing rigid and flexible protein alignments, respectively. 
CE performs a rigid alignment in order to minimize RMSD and FATCAT allows for a constrained number of twists in the protein chain in order to find a more flexible alignment before minimizing RMSD.

Whereas pairwise structure alignments only depend on the sequence of \ca coordinates, protein structure searches often introduce a further dependence on the sequence order of amino acids.
This approach often takes the form of clustering proteins based on sequences and pre-calculating results for pairwise alignments among cluster representatives. 
Then, these pre-calculated results are used for filtering the number of structures used for comparisons against a query protein.
The exact formula for combining the use of representatives and pre-calculated results varies from system to system.
However, all systems using this approach share the same disadvantage, an indirect dependence on amino acid sequences. 
In the absence of a reliance on sequence representatives and pre-calculated results, and without sacrificing accuracy, response times suffer greatly, often taking upwards of an hour for queries to complete.

For protein structure searches, VAST \citep{Gilbrat1996} and the FATCAT server \citep{Ye2004} are among the most popular. 
Nonetheless, these searches are slow in comparison to mTM, SSM, and CATHEDRAL when pre-calculated results are not used. 
If given a known protein domain, VAST can return structural neighbors in seconds using pre-calculated results. 
However, if uploading a PDB file where pre-calculated results are not used, response times for VAST can exceed 30 minutes. 
Similarly, the FATCAT server, that does not use pre-calculated results, can take over an hour to send results for a search against PDB-90 representatives \citep{Prlic2010}.

The PhyreStorm server \citep{Mezulis2016} provides another notable protein structure search.
Here too, it depends on PDB pre-calculated pairwise comparisons among PDB-40 cluster representatives \citep{Prlic2010}. 
Notably, PhyreStorm returns good results within minutes based on an uploaded PDB file.

Given the above, there remains a need for a purely geometric protein structure search.
For the serendipitous exploration of relations between protein structures performed in the trenches, this search should be fast. 
Moreover, with a $~10\%$ yearly growth rate of solved structures deposited in the PDB \citep{gkw1000}, this search should be scalable. 
At a minimum, RUPEE takes a significant step in this direction as will be shown below.

@@

Broadly, we define a linear encoding of protein structure and convert this linear encoding into a bag of features amenable to big data processing.
Protein structure searches that use linear encodings are not unique \cite{Carpentier2005,Daniluk2011,Ritchie2012}.
The novelty of our approach lies in its remarkable performance given its simplicity. 
Additionally, elements of our approach can be isolated and found to be useful in their own right.

@@

@@

\begin{figure*}[tb]
\centering
\includegraphics{torsion_plots}
\caption{Ramachandran plot (right) and polar plot (left) of randomly sampled torsion angles}
\label{fig:torsion_plots}
\end{figure*}

Despite their utility and familiarity, Ramachandran plots \cite{Ramachandran1968} represent angular data using a square plot better suited for scalar data.
This leads to the unwieldy arrangement where the top part of the plot is continuous with the bottom and the left is continuous with the right.

To identify regions of torsion angles, we randomly sampled 10,000 residues from high-resolution, CATH s35 representatives to account for precision and redundancy, respectively. 
A Ramachandran plot of the sampled torsions angles is shown to the left in \figurename~\ref{fig:torsion_plots}. 
As can be seen, a single cluster of residues, consisting primarily of $\beta$-strands, appears at all 4 corners of the Ramachandran plot.

This continuity problem was partially addressed in \cite{Karplus2010} using \emph{wrapped} and \emph{mirrored} plots. 
Both wrapped and mirrored plots take advantage of the sparsely populated areas of the Ramachandran plot at $\phi = 0\degree$ and $\psi = -120\degree$.
However, with larger samples of torsion angles, the area at $\psi = -120\degree$ becomes less sparse. 
The use of a polar plot resolves this elegantly by only requiring one break in continuity at $\phi = 0\degree$.

To the right in \figurename~\ref{fig:torsion_plots}, we plot the same torsion angles appearing in the Ramachandran plot using a polar plot. 
In this plot, $\phi$ corresponds to the radius $r$ and $\psi$ corresponds to the angle $\theta$ in traditional polar plots. 
Notice the residues appearing at the 4 corners of the Ramachandran plot now appear in one continuous region of the polar plot centered at $\phi = \pm180\degree$ and $\psi = \pm180\degree$.

@@

The polar plot described above is used to define torsion angle regions for each secondary structure assignment. 
The eight DSSP secondary structure assignment codes defined in \cite{Kabsch1983} divide into three groups in which torsion angle regions are roughly the same: `G',`H',`I', and `T' corresponding to $3_{10}$-helix, $\alpha$-helix, $\pi$-helix, and turn, respectively; `E' and `B' corresponding to $\beta$-strand and $\beta$-bridge, respectively; and `S' and `C' corresponding to bend and coil, respectively.

@@

\begin{figure*}[tb]
\begin{center}
\includegraphics{torsion_refs}
\end{center}
\caption{Polar plots of randomly sampled torsion angles with designated descriptors for region and DSSP code combinations}
\label{fig:torsion_refs}
\end{figure*}

@@

\begin{table}[t]
    \caption{DSSP Code/Descriptor Map}
    \label{tab:codes}
\centering
\begin{tabular}{lll}
\toprule
    \head{DSSP Codes} & \head{SS} & \head{Descriptors}\\
\midrule
    G, H, I & helix & 1-4\\
    E & strand & 5-7\\
    S, C & bend, coil & 8-10\\
    T & turn & 11\\
    B & bridge & 12\\
\bottomrule
\end{tabular}
\end{table}

Our initial use of Ramachandran plots to identify regions resulted in the frequent case of highly similar domains from the same CATH s95 clusters being considered significantly different. 
Alignment of these domains in PyMOL \cite{PyMOL} and the development of a script to label residues with their descriptors showed that slight differences in structure often resulted in distinct descriptors being assigned to aligned residues. 
Once we defined regions using the less confusing polar plots, this discrepancy was no longer observed.

@@

Secondary structure assignments for termini are not always consistent among the variety of secondary structure assignment tools \cite{Martin2005}. 
We also observed this to be the case for DSSP assignments across homologous domains. 
Here, we introduce a heuristic we call \emph{secondary structure extension} (SSE) to address this variance.

@@

Though \cite{Martin2005} lends some support to this heuristic, we use it primarily based on its performance.

@@

Once a linear encoding for a protein structure is obtained, it needs to be further transformed into a representation suitable for fast and scalable similarity comparisons to other structures.
The processing of text documents within Information Retrieval~(IR) has long been used to satisfy these requirements using bag representations.
There are two distinct categories of representations for documents, syntactic and semantic, and much of the research applying IR to protein structure search has focused on the latter \cite{Aungand2004,Zhang2010,Budowski2010}.

We have adapted the syntactic approach to document similarity, often referred to as shingling \cite{Broder1997a}, to our linear encoding of protein structure. 
We transform a linear sequence of descriptors into a multiset of shingles consisting of 4 consecutive descriptors.
The overlap between shingles ensures some of the order information within the original sequence is preserved in the bag.

@@

Next, each shingle $s$ is hashed to an integer as shown in~(\ref{E:hashdef}). 
The hash function used is a simplification of the hash function used in the Rabin-Karp algorithm \cite{Karp1987}.
The prime number $13$ is used as the base since it is large enough to spread the descriptor values out in hash space without collisions. 
\begin{gather}\label{E:hashdef}
    s_{hash} = s_1 \times 13^3 + s_2 \times 13^2 + s_3 \times 13 + s_4
\end{gather}
Subsequently, the multiset in (\ref{E:shinglebag}) becomes the following bag of integers.
\begin{align}\label{E:hashbag}
    \begin{split}
    \{\,&11900, 11900, 11906, 11984, 12992, \\
        &26096, 25082, 11900, 11900, 11900, \\
        &11900, 11905, 11969 \,\}
    \end{split}
\end{align}

@@

Now that we have a representation of a protein structure as a bag of integers, similarity between any two structures $a$ and $b$ is defined as the Jaccard similarity \cite{Levan1971} for multisets,
\begin{align}
    J(a,b) = \frac{\sum_i min(a_i, b_i)}{\sum_i max(a_i,b_i)}\text{,}
\end{align}
where $i$ ranges over all possible shingle hashes $s_i$ and $a_i$ and $b_i$ give the counts of shingle hash $s_i$ in structures $a$ and $b$, respectively.

@@

In IR, the bag of shingles representation of documents is used in the near dupe clustering of documents \cite{Broder1997b}.
One application of near dupe clustering is in the review stage of Electronic-Discovery \cite{Joshi2008}, which is the most expensive stage in a discovery process. 
Often millions of documents must be examined by a staff of attorneys to make a reasonable effort at providing all documents relevant to the discovery request. 
Grouping documents into near dupe clusters and assigning all documents within a cluster to a single reviewer reduces duplication of effort.

In the case of near dupe clustering, each document must be compared to every other document in the collection, taking quadratic time.
For this task, min-hashing \cite{Broder1998} and locality sensitive hashing (LSH) \cite{Indyk1998} can be combined to reduce this to subquadratic time. 
Although we do not near dupe cluster domains, we can still leverage the techniques of min-hashing and LSH to speed up protein structure search by a large constant factor.

Min-hashing is used to randomly select items from a set of items by repeatedly randomly hashing the items, sorting the hashes into a list, and then selecting the minimum item in each permuted list.
If the same random permutation of items is performed on each set of items in a collection, the key result is that the probability of matching min-hashes is equal to the Jaccard similarity \cite{Broder1998}.
In order to approximate the Jaccard similarity for a given pair of sets, a sufficient number of min-hashes must be obtained.

In our case, the items are bags of shingle hashes for protein structures from which we obtain 60 min-hashes as described in \cite{Rajaraman2012}. 
Given the key result above, the Jaccard similarity can now be approximated by the proportion of matching min-hashes.

Weighted min-hashing \cite{Chum2008} with integral weights is an extension to min-hashing that can be applied to multisets where the frequency of a hash within a multiset is used as its weight.

Next, we use the LSH banding technique as described in \cite{Rajaraman2012}.
The key result of the banding technique is that if \emph{any} band positions are a match for a given pair of structures, the probability that a specific similarity threshold has been met can be calculated. 
We use 20 bands of 3-min-hashes where the probability of a Jaccard similarity of 60\% or greater is approximately 99\%. 
Banding allows the problem of finding similar items to be parallelized across bands since all that is needed for a match is a single band match.
